---
title: "Reproducible Research: Peer Assessment 1"
author: "Michael Wood"
date: "January 21, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)

# Create data directory if necessary                                            
if(!file.exists("./data")){dir.create("./data")}                                
                                                                                
# Set fileURL and download data if necessary                                    
fileURL <- ""                                                                   
fileDL <- "./data/activity.csv"                                                 
if(!file.exists(fileDL)){                                                       
        download.file(fileURL, destfile=fileDL, method="curl")                  
} 
```
## Pear-graded Assignment:  Course Project 1
### Code for reading in the dataset and/or processing the data
```{r Question 1}
# Read CSV file
fileDL <- "./data/activity.csv"
csvData <- read.csv(fileDL)

# Maintain CSV data in imported form 
rawData <- csvData

# Remove NA data from data set
data.NArm <- subset(rawData, steps != "NA")                                     
data.NArm$date <- as.Date(as.character(data.NArm$date))
```

### Histogram of the total number of steps taken each day
```{r Question 2}
# Sum total steps by date
sumByDate <- as.data.frame(tapply(data.NArm$steps, data.NArm$date, FUN=sum))    
colnames(sumByDate) <- c("TotalSteps")
sumByDate$TotalSteps <- as.numeric(as.character(sumByDate$TotalSteps))
```
```{r Question 2 Plot}
# Create histogram showing total steps by day
hist(sumByDate$TotalSteps)
```


### Mean and median number of steps taken each day
```{r Question 3}
# Create data frame containing mean steps by date
meanByDate2 <- as.data.frame(tapply(data.NArm$steps, data.NArm$date, FUN=mean)) 
colnames(meanByDate2) <- c("MeanSteps")         

# Create data frame containing median steps by date
medianByDate2 <- as.data.frame(tapply(data.NArm$steps, data.NArm$date, FUN=median)) 
colnames(medianByDate2) <- c("MedianSteps")
```


### Time series plot of the average number of steps taken
``` {r Question 4}
# Create data frame containing average number of steps over time
meanByInterval <- tapply(data.NArm$steps, data.NArm$interval, FUN=mean)         
meanByInterval <- as.data.frame(cbind(meanByInterval, row.names(meanByInterval)))
colnames(meanByInterval) <- c("MeanSteps", "Interval")                          
meanByInterval$MeanSteps <- as.numeric(as.character(meanByInterval$MeanSteps))  
meanByInterval$Interval <- as.integer(as.character(meanByInterval$Interval))    
```
``` {r Question 4 Plot}
qplot(Interval, MeanSteps, data = meanByInterval, geom="line")+ theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


### The 5-minute interval that, on average, contains the maximum number of steps
``` {r Question 5}
subset(meanByInterval, MeanSteps == max(meanByInterval$MeanSteps))
```


### Code to describe and show a strategy for imputing missing data
``` {r Question 6}
na.sum <- sum(is.na(rawData))
```
There are `r na.sum` NA values in the original data set.

One solution to this problem is to replace missing values with average values based on the correlating interval.
``` {r Question 6 Part II}
# Use mean for given 5-minute interval
# Create new data set replacing NA values with mean by interval
completeData <- rawData                                                         
for( i in 1:nrow(completeData)) {                                               
        if( is.na(completeData[i, "steps"])) {                                  
                completeData[i, ]$steps <- meanByInterval[meanByInterval$Interval==completeData[i,"interval"], "MeanSteps"]
        }                                                                       
}
completeData$date <- as.Date(as.character(completeData$date))                   
sumByDate3 <- as.data.frame(tapply(completeData$steps, completeData$date, FUN=sum))
colnames(sumByDate3) <- c("TotalSteps")                                         
```


### Histogram of the total number of steps taken each day after missing values are imputed
The original summary values:
``` {r Question 7}
# Summaries showing variations in data sets coming from replacing values with mean values.
summary(sumByDate)  
```
The summary values of the repaired data set:
``` {r Question 7 Part II}
summary(sumByDate3)
```
Clearly the min, max, median, and mean values varied little or not at all, but the quartile values changed by 500-1000 steps.
``` {r Question 7 Plot}
# Generate histogram of repaired data set
hist(sumByDate3$TotalSteps)
```


### Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
``` {r Question 8}
# Create data frame containing date and weekday information
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
DoW <- as.data.frame(weekdays(completeData$date))

# Generate new weekday/weekend column based on correlated days of the week
for( i in 1:nrow(DoW)) {                                                        
        if( DoW[i,1] %in% weekdays ) {                                          
#               print("DAY")                                                    
                DoW[i,2] <- as.character("weekday")                             
        } else {                                                                
#               print("END")                                                    
                DoW[i,2] <- as.character("weekend")                             
        }                                                                       
}                                                                               
                                                                                
completeData2 <- cbind(completeData, DoW[,2])                                   
colnames(completeData2) <- c("Steps", "Date", "Interval", "WeekDayEnd")

# Generate mean values based on weekday or weekend
weekdayData <- completeData2[completeData2$WeekDayEnd == "weekday",]            
weekdayDataMean <- as.data.frame(tapply(weekdayData$Steps, weekdayData$Interval, FUN=mean))
colnames(weekdayDataMean) <- c("MeanSteps")                                     
weekdayDataMean.df <- as.data.frame(cbind(weekdayDataMean$MeanSteps, row.names(weekdayDataMean), rep_len("weekday", length.out=nrow(weekdayDataMean))))
                                                                                
weekendData <- completeData2[completeData2$WeekDayEnd == "weekend",]            
weekendDataMean <- as.data.frame(tapply(weekendData$Steps, weekendData$Interval, FUN=mean))
colnames(weekendDataMean) <- c("MeanSteps")                                     
weekendDataMean.df <- as.data.frame(cbind(weekendDataMean$MeanSteps, row.names(weekendDataMean), rep_len("weekend", length.out=nrow(weekendDataMean))))
                                                  
# Concatenate weekday/weekend datasets                               
meanData.df <- rbind(weekdayDataMean.df, weekendDataMean.df)                    
colnames(meanData.df) <- c("MeanSteps", "Interval", "WeekDayEnd")               
meanData.df$Interval <- as.integer(as.character(meanData.df$Interval))          
meanData.df$MeanSteps <- as.numeric(as.character(meanData.df$MeanSteps))        
                                       
# Generate panel plot comparing weekday to weekend step averages                                         
qplot(Interval, MeanSteps, data = meanData.df, facets = WeekDayEnd~., geom="line")
```